---
title: "Branch length summary stats analysis"
output: 
  prettydoc::html_pretty:
    theme: "cayman"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning=FALSE, message=FALSE, echo= TRUE, include = TRUE)
library(tidyverse)
library(prettydoc)
```

```{r source utils.R, include=FALSE}
sys.source("utils.R", envir = knitr::knit_global())
```

## Goal
The primary goal of this analysis is to determine whether branch length summary statistics differ across protein models. This will be done mainly with code that visualizes these statistics. 

```{r read in the data, include=FALSE}
path_to_data <- "./"

birds_file <- file.path(path_to_data, "bird_empirical_branch_lengths.csv")
mammals_file <- file.path(path_to_data, "mammal_empirical_branch_lengths.csv")
enzymes_file <- file.path(path_to_data, "enzyme_empirical_branch_lengths.csv")



mammals <- read_csv(mammals_file)
enzymes <- read_csv(enzymes_file)
birds   <- read_csv(birds_file)

birds$id=as.character(birds$id)
mega_empirical_dataset<-bind_rows(enzymes, mammals, birds)

```

### First Step: Create Smaller Datasets
The datasets are too large, so let's make smaller datasets comprized of three of each id from each dataset
```{r make mini datasets to work with}
mammals%>%
  filter(id%in%c("ENSG00000000003_AA","ENSG00000000005_AA", "ENSG00000000457_AA"))->mammals_3_id

birds%>%
  filter(id%in%c("100", "1000", "10003"))->birds_3_id

enzymes%>%
  filter(id%in%c("13PK_A", "1A16_A", "1CL1_A"))->enzymes_3_id

```

## Analysis
Okay, now that we have our fun mini-datasets, let's begin to see how the branch length estimates look for each!

```{r summarize branch lengths, include=FALSE}
summarize_branch_lengths(mammals_3_id)->mammals_3_bl
summarize_branch_lengths(enzymes_3_id)->enzymes_3_bl
summarize_branch_lengths(birds_3_id)->birds_3_bl

mega_mini_bl<-bind_rows(mammals_3_bl,enzymes_3_bl,birds_3_bl)

mega_mini_bl%>%
  mutate(dataset= case_when(id%in%c("ENSG00000000003_AA", "ENSG00000000005_AA", "ENSG00000000457_AA") ~"mammals",
                            id%in%c("1CL1_A", "1A16_A", "13PK_A") ~"enzymes",
                            id%in%c("1000","100", "10003") ~"birds"))-> mega_mini_bl2

```


```{r define function for bl measurements, include=FALSE}

Violin_bl_measurements<-function(bl_df, measurement, y_axis_title, plot_title)
{
bl_df%>%
  select({{measurement}}, model, id, ASRV, dataset)%>%
  filter(ASRV=="TRUE")%>%
  group_by(model, id, dataset)%>%
  ggplot(aes(x=model, y={{measurement}}, fill=model))+
  geom_violin()+
  geom_point()+
  stat_summary()+
  scale_fill_brewer(palette = "Dark2")+
  labs(x="Model", y=y_axis_title, title=plot_title)
}

```

```{r vizualize branch lengths}


Violin_bl_measurements(mega_mini_bl2, treelength, "Treelength", "Estimated Treelengths for Selected Id's")
Violin_bl_measurements(mega_mini_bl2, mean_bl, "Mean_bl", "Estimated Mean Branchlengths for Selected Id's")
Violin_bl_measurements(mega_mini_bl2, max_bl, "Max_bl", "Estimated Maximum Branchlengths for Selected Id's")

```
<br><br>
From this, we see that FLU appears to overestimate the branchlength-related measurements while Poisson underestimates!
<br><br>

### Now, Let's address our goal by utilizing linear models to determine if branch length summary statistics do indeed differ across protein models.

```{r, linear model statistic~ protein_model + ASRV}
linear_model_function_curlies<-function(input_branchlength_df, dependent_variable_column)
{
  input_branchlength_df %>%
    mutate(ASRV_modified= if_else(ASRV==TRUE, "Yes", "No")) -> df_for_modeling #need to make ASRV non-boolean
  
  lm({{dependent_variable_column}} ~ model+ dataset+ ASRV_modified, data= df_for_modeling) %>% #since we are referring to a column,we need to use curlies
    aov() %>%
    TukeyHSD()->fitted_model
  
  fitted_model$model %>% 
    as_tibble(rownames = "Compare") #this will make the output a little nicer
}
  
linear_model_function_curlies(mega_mini_bl2, mega_mini_bl$mean_bl)
linear_model_function_curlies(mega_mini_bl2, mega_mini_bl$treelength)
linear_model_function_curlies(mega_mini_bl2, mega_mini_bl$max_bl)
```
<br><br>



