---
title: "Branch length summary stats analysis"
output: 
  prettydoc::html_pretty:
    theme: "cayman"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning=FALSE, message=FALSE, echo = TRUE, include = TRUE)
library(tidyverse)
library(prettydoc)
```

```{r source utils.R, include=FALSE}
source("utils.R", local = knitr::knit_global())
```

## Goal
The primary goal of this analysis is to determine whether branch length summary statistics differ across protein models. This will be done mainly with code that visualizes these statistics. 

```{r read in the data, include=FALSE}
path_to_data <- "./"

birds_file <- file.path(path_to_data, "bird_empirical_branch_lengths.csv")
mammals_file <- file.path(path_to_data, "mammal_empirical_branch_lengths.csv")
enzymes_file <- file.path(path_to_data, "enzyme_empirical_branch_lengths.csv")



mammals <- read_csv(mammals_file)
enzymes <- read_csv(enzymes_file)
birds   <- read_csv(birds_file)

birds$id=as.character(birds$id)
mega_empirical_dataset<-bind_rows(enzymes, mammals, birds)

```

The datasets are too large, so let's make smaller datasets comprized of three of each id from each dataset
```{r make mini datasets to work with}
mammals%>%
  filter(id%in%c("ENSG00000000003_AA","ENSG00000000005_AA", "ENSG00000000457_AA"))->mammals_3_id

birds%>%
  filter(id%in%c("100", "1000", "1003"))->birds_3_id

enzymes%>%
  filter(id%in%c("13PK_A", "1A16_A", "1CL1_A"))->enzymes_3_id

print(birds_3_id)
print(enzymes_3_id)
print(mammals_3_id)
```

## Analysis
Okay, now that we have our fun mini-datasets (and the utils script has been sourced), let's begin to see how the branch length estimates look for each!

```{r analyze branch lengths, include=TRUE}
```
