---
title: "Actual Branch Length Analysis"
output: 
  prettydoc::html_pretty:
    theme: "cayman"
params: 
  mini: FALSE
  mini_n: 30
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, include = TRUE)
library(tidyverse)
library(broom)
path_to_data <- "./" 
```

## Goal
The primary goal of this analysis is to determine if the literal branch length estimates themselves differ across protein models. 
<br>
First, let's read in the datasets and create smaller versions of them that are easier to work with, just like last time.

```{r read in the data and make smaller datasets, message=FALSE}

birds_file <- file.path(path_to_data, "bird_empirical_branch_lengths.csv")
mammals_file <- file.path(path_to_data, "mammal_empirical_branch_lengths.csv")
enzymes_file <- file.path(path_to_data, "enzyme_empirical_branch_lengths.csv")


birds   <- read_csv(birds_file) %>% mutate(dataset =  "birds", id = as.character(id)) #might as well mutate here so we can bind rows later 
mammals <- read_csv(mammals_file) %>% mutate(dataset =  "mammals", id = as.character(id))
enzymes <- read_csv(enzymes_file) %>% mutate(dataset =  "enzymes", id = as.character(id))

# Downsample for mini analysis only
minify <- function(df) 
{
  df %>%
    group_by(id) %>% 
    nest() %>% #condense each id into its own tibble-like thing 
    ungroup() %>%
    slice(1:params$mini_n) %>% #this takes the first *however many* rows from the dataset. The *however many* is specified by the params$mini_n
    unnest(cols = c(data))
}

if (params$mini == TRUE) #if we want to use the mini versions of the datasets, then use the mini function on each of them!!
{
  minify(birds) -> birds 
  minify(enzymes) -> enzymes
  minify(mammals) -> mammals
}


bind_rows(mammals, enzymes, birds) -> full_data #this is just one big data frame consisting of the three datasets


```

## The enzyme, mammal, and enzyme data all together

```{r show_bl, echo=TRUE}
print(full_data) 
```



## The data pivoted wide to show all models' branch lengths and pivoted again to compare Poisson branch lengths with all others 
```{r, pivot full data, echo=TRUE}
  
full_data %>%
  unite(combined_model, model, ASRV) %>%
  group_by(id, dataset) %>%
  pivot_wider(names_from = combined_model, 
              values_from = branch_length) %>%
  ungroup() -> input_data

input_data%>%
pivot_longer( -c(id, node, dataset, Poisson_FALSE), names_to = "model_type", values_to = "branchlength")->input_poisson_f

```


## Let's see how one looks to get a better sense of the data we're modeling

```{r a_scatter}
input_poisson_f %>%
  filter(id == "13PK_A") %>%
  ggplot(aes(x = Poisson_FALSE, 
             y = branchlength, 
             color = model_type)) + 
  geom_point() + 
  geom_smooth(method = "lm", color = "black") +
  geom_abline(color = "red") + 
  facet_wrap(vars(model_type)) +
  theme(legend.position = "none")
```


```{r,show input data}
print(input_data)
print(input_poisson_f)
```


## Perform a glm of OFFSETTED branchlength~Poisson_FALSE, with the help of `tidyr::nest()` and `{broom}`.

> The slope is _forced to be 1_. The intercept therefore tells us the _bias of the esimator_ compared to truth.
> We "define truth" as Poisson_FALSE
> Intercept means: What is the difference between other model's bl - Poisson_FALSE bl

```{r, pivot full data and use lm, echo=TRUE}
input_poisson_f%>%
  select(-node) %>%
  group_by(id, dataset, model_type)%>%
  tidyr::nest()%>%
  mutate(lm_fit=map(data, ~glm(branchlength ~ offset(Poisson_FALSE), data=.)))->nested


nested%>%
  select(-data)%>%
  mutate(tidied=map(lm_fit, broom::tidy))%>%
  unnest(cols = c(tidied))->unnested

unnested%>%
  filter(term == "(Intercept)")%>%
  select(id, dataset, model_type, estimate)%>%
  rename(intercept=estimate)->Intercepts_df

readr::write_csv(Intercepts_df, "biasintercepts_poisson_false.csv")
```

## Pull out the intercepts from the lm 
```{r, show intercepts dataframe}
#print(Intercepts_df)

```

## Create a jitter of the intercepts 
```{r, ggplot the Intercepts, fig.width=10, fig.height=8}
#ggplot(Intercepts_df)+
#  aes(x=model_type, y=intercept, fill=model_type)+
#  geom_jitter(width = 0.1, pch = 21) + 
#  theme_classic()+
#  scale_fill_brewer(palette = "Dark2", name= "Model Type")+
#  labs(x="Model Type", 
#       y= "Intercept of lm branch length ~ offset(Poisson_FALSE)", title = "Jitter of branch length lm intercepts")
```

## Analysis

-intercepts of this lm represent the Poisson_FALSE branch length when the other models' branch length is zero (y-intercept occurs when x=0, and in our case, the other models' branch lengths serve as our predictor)

-There is less deiscepency bewteen ASRV- plots, but more when you compare ASRV- to ASRV+

- It appears that Poisson_False actually slightly overestimates the FLU_TRUE, JTT_TRUE, and LG_TRUE branch lengths (when these predictors are at zero, Poisson_FALSE estimates the branch length to be greater than zero)
