---
title: "Practice Plots"
subtitle: "3/1/2022"
author: "Courtney Charlesworth and Stephanie Spielman"
output: 
  html_document:
    theme: readable
    highlight: pygments
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)

#path to data file
path_to_ic <- file.path(here::here(), "results_now", "simulation_ic_ranks_weights_per_bl.csv")

#read data file
ic_data <-read_csv(path_to_ic) %>% 
  rename(np_sim_model = site) %>%
  mutate(model = factor(model, levels = c("FLU", "LG", "JTT", "WAG", "Poisson")))

```

<br>

### Variables

+ `np_sim_model` = nucleoprotein simulation model

+ `sim_branch_length` = amount of branch length provided to simulator

+ `model` = the substitution models (FLU, WAG, JTT, LG, Poisson = equal)

+ `ASRV` = gamma distribution (yes or no)

  - rate variation/heterogeneity, fancy statistics  

+ `ic_type` = information criterion (tells us which model is better or worse)

  - there are 3 that we're looking at: AIC, AICc, and BIC

+ `ic_rank` = ranks model 1-10 (1 is best, 10 is worst)

+ `ic_weight` = shows magnitude of difference between how good models did

  - there can be big differences and 1-10 (`ic_rank`) doesn't properly show this

  - sums to 1 (best), 0 is really bad
  
  
*goal: which model is best and in what conditions?*

<br>

### Plot 1
```{r plot1}
#how many times the models had a #1 ranking
ic_data %>%
  filter(ic_rank == 1) %>%
  count(model, name = "ic_rank_1_count") %>%
  #LG has 0 #1 rankings so it got filtered out
  add_row(model = "LG", ic_rank_1_count = 0) %>% 
  print() %>%
  ggplot() +
  aes(x = model,
      y = ic_rank_1_count) +
  geom_col(fill = "turquoise4") +
  #text of count because WAG and Poisson so small they don't show up
  geom_text(aes(x = model,
                y = ic_rank_1_count + 2000,
                label = ic_rank_1_count))

```

<br>

### Plot 2
```{r plot2}
#really ugly plot. just wanted to look at everything at once
ic_data %>%
  ggplot() +
    aes(x = ic_type,
        y = ic_weight) +
    geom_col() +
    facet_grid(cols = vars(model),
               rows = vars(ASRV),
               scales = "free_y")

```

<br>

### Plot 3
```{r plot3}
#distribution
ic_data %>%
  ggplot() +
    aes(x = ic_rank) +
    geom_histogram(bins = 5,
                   fill = "blue") +
    facet_grid(cols = vars(model),
               rows = vars(ic_type))

```

IC thinks Poisson sucks 

