---
title: "Practice Plots"
subtitle: "1/22/2022"
author: "Courtney Charlesworth"
output: 
  html_document:
    theme: readable
    highlight: pygments
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(dplyr)

data <- read_csv("copy_simulation_branch_lengths_counts.csv")

```

## which model (WAG, Poisson, etc.) best predicts the number of substitutions/site (and in what conditions is one better than the other)

<br>

+ substitutions = insertion/deletion events = mutations
+ substitutions/site = branch length

+ branch length = the output of the exp 

+ truth = simulation = `persite_count`
+ inferred = what the model (WAG, Poisson, etc.) thinks sub/site was  = `branch_length`

<br>

## Variables
simulation

+ `count` = total number of amino acid changes in the simulation   
+ `site` = #1-498 flu nucleoprotein (category)
+ `sim_branch_length` = total amount of branch length
+ `persite_count` = count/10,000 = truth (how many substitutions)

model

+ `model` = the substitution models (FLU, WAG, JTT, LG, Poisson = equal)
+ `branch_length` = how many substitutions the models think happened     
+ `ASRV` = gamma distribution (yes or no) = rate variation/heterogeneity 

<br>

### Plot 1

```{r plot1}
#sim_branch_length vs.branch_length; bl vs. bl
data %>%
  ggplot() +
  aes(x = branch_length, #range = 0-30 
      y = sim_branch_length) + #range = 0-3, these numbers don't look random
  geom_point()

```

<br>

### Plot 2

```{r plot2}
#truth vs. inferred (persite_count vs. branch_length)
data %>%
  ggplot() +
  aes(x = branch_length, #inferred
      y = persite_count) + #truth (0.01 - 3.00); after 3.00, it's a new site and starts back at 0.01
  geom_point(alpha = 0.5)

```

count (number of substitutions) increases as branch length (sim_branch_length) increases

longer branch lengths = more simulations can be run 

<br>

### Plot 3

```{r plot3}
#truth vs. inferred (persite_count vs. branch_length) per model
#same plot as above but faceted by model
data %>%
  ggplot() +
  aes(x = branch_length, #inferred
      y = persite_count) + #truth
  geom_point(alpha = 0.5) +
  facet_wrap(vars(model),
             scales = "free")

```

Poisson has shortest branch_length values

makes sense because it assumes everything is equal

<br>

### Plot 4
```{r plot4A}
#bar plot of branch length faceted by model, comparing ASRV
data %>%
  filter(site == 1,
         sim_branch_length == 1.40) %>%
  print() %>%
    ggplot() +
      aes(x = ASRV,
          y = branch_length, #inferred
          fill = ASRV) + 
      geom_col() +
      facet_wrap(vars(model))

```


```{r plot4B}
#same as above; different part in the data
data %>%
  filter(site == 290,
         sim_branch_length == 2.30) %>%
  print() %>%
    ggplot() +
      aes(x = ASRV,
          y = branch_length, #inferred
          fill = ASRV) + 
      geom_col() +
      facet_wrap(vars(model))

```

+G seems like it tends to be a little higher

largest differences seem to be in FLU and LG, smaller differences in JTT and WAG, Poisson the same for both

```{r plot4C}
#mean branch length for each model/ASRV combination
data %>%
  group_by(model, ASRV) %>%
  summarize(mean_branch_length = mean(branch_length)) %>%
  print() %>%
    ggplot() +
      aes(x = ASRV,
        y = mean_branch_length, #inferred
        fill = ASRV) + 
        geom_col() +
        facet_wrap(vars(model))

```

+G tends to be a little higher

largest differences seem to be in FLU and LG, smaller differences in WAG then JTT, Poisson still basically the same

<br>

### Plot 5
```{r plot5}
#total branch length vs. # of substitutions
data %>%
  ggplot() +
  aes(x = persite_count, #truth
      y = sim_branch_length) + 
  geom_point(alpha = 0.5)

```

positive relationship

### Plot 6
```{r plot6A}
#density plot of truth column
data %>%
  ggplot() +
  aes(x = persite_count, #truth
      fill = model) +
  geom_density() +
  facet_wrap(vars(model))

```

everything looks the same (expected, can see that the numbers are the same for each model)

```{r plot6B}
#density plot of inferred column
data %>%
  ggplot() +
  aes(x = branch_length, #inferred
      fill = model) + 
  geom_density() +
  facet_wrap(vars(model),
             scales = "free") 

```

different distributions


## Questions

+ Is `sim_branch_length` given to the simulation? and then output is count?
+ `sim_branch_length` can be stated as every site has had _____ changes on average?


+ `persite_count` = count/10,000 or 1e4 sites. where is 10,000 coming from?
+ "Each individual simulation uses a single site parameterization from NP and 1e4 sites. Total of 498 sites"
+ i think the usage of sites in maybe 2-3 contexts is confusing me


+ why would +G make the branch_length longer?
+


