---
title: "Practice Plots"
subtitle: "1/22/2022"
author: "Courtney Charlesworth and Stephanie Spielman"
output: 
  html_document:
    theme: readable
    highlight: pygments
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse) # no need to also load dplyr b/c dplyr is loaded with tidyverse

# use reproducible paths
path_to_data <- file.path(here::here(), "results", "simulation_branch_lengths_counts.csv")

# avoid getting confused with poor use of "site" term all over the place
data <- read_csv(path_to_data) %>% rename(np_sim_model = site)
```

## which model (WAG, Poisson, etc.) best predicts the number of substitutions/site (and in what conditions is one better than the other)

<br>

+ substitutions = insertion/deletion events = mutations
+ substitutions/site = branch length

+ branch length = the output of the exp 

+ truth = simulation = `persite_count`
+ inferred = what the model (WAG, Poisson, etc.) thinks sub/site was  = `branch_length`

<br>

## Variables
simulation

+ `count` = total number of amino acid changes in the simulation.   
+ `site` = #1-498 flu nucleoprotein (category). "Which simulation model was used."
+ `sim_branch_length` = total amount of branch length provided to simulator. "Which simulation branch length"
+ `persite_count` = count/10,000 = truth (how many substitutions)

model

+ `model` = the substitution models (FLU, WAG, JTT, LG, Poisson = equal)
+ `branch_length` = how many substitutions the models think happened     
+ `ASRV` = gamma distribution (yes or no) = rate variation/heterogeneity 

<br>



### Plot 1

> Not a plot for the shiny app 

```{r plot1}
#sim_branch_length vs.branch_length; bl vs. bl
data %>%
  ggplot() +
  aes(x = branch_length, #range = 0-30 
      y = sim_branch_length) + #range = 0-3, these numbers don't look random
  geom_point() + 
  geom_abline(color = "red")

```


Each individual simulation has a total tree length of X (where X = `sim_branch_length`) where $X \in {0.01, 0.05, seq(0.1, 3, 0.1)}$

<br>

### Plot 2

```{r plot2}
#truth vs. inferred (persite_count vs. branch_length)
data %>%
  ggplot() +
  aes(y = branch_length, #inferred
      x = persite_count) + #truth (0.01 - 3.00); after 3.00, it's a new site and starts back at 0.01
  geom_point(alpha = 0.5)+ 
  geom_abline(color = "red", size = 1) +
  labs(x = "(TRUTH) amino acid changes per site IN SIMULATION", 
       y = "(INFERRED) MODEL GUESSES amino acid changes per site") + 
  # Zoom into only simulations with very FEW changes
  lims(x = c(0, 0.1), y = c(0, 0.1)) + 
  # Highlight different results across difference INFERENCE conditions
  facet_grid(rows = vars(model),
             cols = vars(ASRV))

```

count (number of substitutions) increases as branch length (sim_branch_length) increases

longer branch lengths = more simulations can be run 

<br>

### Plot 3

```{r plot3}
#truth vs. inferred (persite_count vs. branch_length) per model
#same plot as above but faceted by model
data %>%
  ggplot() +
  aes(x = branch_length, #inferred
      y = persite_count) + #truth
  geom_point(alpha = 0.5) +
  facet_wrap(vars(model),
             scales = "free")

```

Poisson has shortest branch_length values

makes sense because it assumes everything is equal

<br>

### Plot 4
```{r plot4A}
#bar plot of branch length faceted by model, comparing ASRV
data %>%
  filter(site == 1,
         sim_branch_length == 1.40) %>%
  print() %>%
    ggplot() +
      aes(x = ASRV,
          y = branch_length, #inferred
          fill = ASRV) + 
      geom_col() +
      facet_wrap(vars(model))

```


```{r plot4B}
#same as above; different part in the data
data %>%
  filter(site == 290,
         sim_branch_length == 2.30) %>%
  print() %>%
    ggplot() +
      aes(x = ASRV,
          y = branch_length, #inferred
          fill = ASRV) + 
      geom_col() +
      facet_wrap(vars(model))

```

+G seems like it tends to be a little higher

largest differences seem to be in FLU and LG, smaller differences in JTT and WAG, Poisson the same for both

```{r plot4C}
#mean branch length for each model/ASRV combination
data %>%
  group_by(model, ASRV) %>%
  summarize(mean_branch_length = mean(branch_length)) %>%
  print() %>%
    ggplot() +
      aes(x = ASRV,
        y = mean_branch_length, #inferred
        fill = ASRV) + 
        geom_col() +
        facet_wrap(vars(model))

```

+G tends to be a little higher

largest differences seem to be in FLU and LG, smaller differences in WAG then JTT, Poisson still basically the same

<br>

### Plot 5
```{r plot5}
#total branch length vs. # of substitutions
data %>%
  ggplot() +
  aes(x = persite_count, #truth
      y = sim_branch_length) + 
  geom_point(alpha = 0.5)

```

positive relationship

### Plot 6
```{r plot6A}
#density plot of truth column
data %>%
  ggplot() +
  aes(x = persite_count, #truth
      fill = model) +
  geom_density() +
  facet_wrap(vars(model))

```

everything looks the same (expected, can see that the numbers are the same for each model)

```{r plot6B}
#density plot of inferred column
data %>%
  ggplot() +
  aes(x = branch_length, #inferred
      fill = model) + 
  geom_density() +
  facet_wrap(vars(model),
             scales = "free") 

```

different distributions


## Questions

+ Is `sim_branch_length` given to the simulation? and then output is count? SJS: YES!!
+ `sim_branch_length` can be stated as every site has had _____ changes on average? yes...every site has had ___ NEUTRAL CODON changes on average. This is something you don't want to care about. 

+ `sim_branch_length` is *proportional* to the amount of evolutionary change in the given simulation. 


+ `persite_count` = count/10,000 or 1e4 sites. where is 10,000 coming from?
+ "Each individual simulation uses a single site parameterization from NP and 1e4 sites. Total of 498 sites"
+ i think the usage of sites in maybe 2-3 contexts is confusing me


+ why would +G make the branch_length longer?



## Paired programming plots

This plot should be in the shiny app with user ability to change the site (AKA NP MODEL - YOU MAY NEED TO DEBUG THIS CODE!)
```{r paired1}
data %>%
  filter(site == 3) %>% # Users can change the "site" (NP)
  ggplot() + 
  aes(x = persite_count, 
      y = branch_length) + 
  geom_point() +
    facet_grid(cols = vars(model),
             rows = vars(ASRV)) + 
  geom_abline(color = "red")  + # equality
  geom_smooth(method = "lm", color = "blue", size = 0.5)
```
Each point is a simulation. Each panel is how that simulation was MEASURED under different modeling circumstances.