---
title: "Practice Plots"
subtitle: "2/2/2022"
author: "Courtney Charlesworth and Stephanie Spielman"
output: 
  html_document:
    theme: readable
    highlight: pygments
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)

#path to data file
path_to_bs <- file.path(here::here(), "results", "simulation_bias_slope.csv")

#read data file
bs_data <-read_csv(path_to_bs) %>% 
  select(-slope_p_value, -bias_p_value) %>% 
  rename(np_sim_model = site)

```

### Variables

<br>

+ `np_sim_model` = nucleoprotein simulation model

+ `ASRV` = gamma distribution (yes or no) = rate variation/heterogeneity

+ `model` = the substitution models (FLU, WAG, JTT, LG, Poisson = equal)

+ `slope_when_yint0` = make y int = 0, what is slope value? 1 = :D, <1 less than :D, >1 greater than :D

+ `slope_p_value_corrected` = indicates: is the slope inferred significantly different from the null? 

+ `bias` = if fix slope to 1 (true, expected relationship) what is the y int? pos = higher, neg = inference lower than it should be

+ `bias_p_value_corrected` = indicates: is bias significantly different from 0?

<br>

### Plot 1
```{r plot1}
#scatterplot of bias vs slope_when_yint0
bs_data %>%
  ggplot() +
  aes(x = slope_when_yint0,
      y = bias) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm",
              color = "red",
              size = 0.5)

```

positive relationship, most points concentrated on left side of graph

<br>

### Plot 2
```{r plot2}
#scatterplot of bias vs slope_when_yint0 faceted by model and ASRV
bs_data %>%
  ggplot() +
  aes(x = slope_when_yint0,
      y = bias) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm",
              color = "red",
              size = 0.5) +
  facet_grid(cols = vars(model),
             rows = vars(ASRV),
             scales = "free")

```

interesting differences between the models

<br>

### Plot 3
```{r plot3a}
#density bias
bs_data %>%
  ggplot() +
  aes(x = bias) +
  geom_density(fill = "red") 

```


```{r plot3b}
#density bias_p_value_corrected
bs_data %>%
  ggplot() +
  aes(x = bias_p_value_corrected) +
  geom_density(fill = "red")

```

<br>

### Plot 4
```{r plot4a}
#density slope_when_yint0
bs_data %>%
  ggplot() +
  aes(x = slope_when_yint0) +
  geom_density(fill = "red")

```

```{r plot4b}
#slope_p_value_corrected
bs_data %>%
  ggplot() +
  aes(x = slope_p_value_corrected) +
  geom_density(fill = "red")

```

<br>

### Plot 5
```{r plot5}


```


